\section{Prompting}
Prompting is the process of designing inputs to LLMs in order to obtain the desired outputs.
Effective prompting is essential when building AI systems because it directly affects output quality, reliability, and cost.

In this course, we focus on prompting through \textbf{API calls} rather than consumer user interfaces.
The two settings differ significantly: 
\begin{itemize}
    \item \textbf{User Interfaces}: These are designed for end-users and often include features that help guide the user in crafting effective prompts, such as templates, examples, and interactive feedback.
    They usually don't allow you to access the model directly, so you have less control over the prompting process, but the interface itself can help users get better results.
    \item \textbf{API Calls}: When using APIs, you have more control and flexibility over the prompting process, 
    as you are talking directly to the model. You have to design the prompt yourself, and you can also manage the conversation state and context more effectively.
    Modern APIs may also expose built-in tools such as web search, code execution, and retrieval, which can further improve results.
\end{itemize}
In short, API prompting is developer-centric, while user interfaces are user-centric.
For instance, obtaining structured output can be difficult in many interfaces but straightforward through an API call.

\subsection{Anatomy of a Prompt}
A prompt typically consists of the following components:
\begin{itemize}
    \item \textbf{System Prompt}: Sets the persona, rules and constraints the model follows throughout the conversation (job description).
    Conventionally, the system prompt is the same across all calls, but needs to be repeated since the API is stateless.
    \item \textbf{User Prompt}: The actual task or question you want the model to address. 
    It can include the input data, examples and instructions on how to format the output.
    Conventionally, the user prompt is the part that changes across calls, depending on the specific task or input data.
    \item \textbf{Parameters}: These are settings that influence the model's behavior, such as temperature, top-p, max tokens, and stop sequences.
\end{itemize}

\subsection{API Parameters}
When making API calls to an LLM, you can specify parameters that influence model behavior:
\begin{itemize}
    \item \textbf{Temperature}: Controls output randomness. Higher values (e.g., 0.8) usually produce more diverse and creative text, while lower values (e.g., 0.2) usually produce more focused and consistent text. In practice, the model samples from token probabilities, and temperature shapes that sampling.
    \item \textbf{Top-p}: Controls diversity by sampling only from the smallest set of tokens whose cumulative probability reaches \texttt{p}. It is typically used as an alternative to temperature; in many cases, tuning one is enough.
    \item \textbf{Max Tokens}: Caps response length, helping control both cost and verbosity.
    \item \textbf{Stop Sequences}: Define one or more strings that cause generation to stop when encountered.
    \item \textbf{Frequency Penalty}: Penalizes new tokens based on their existing frequency in the text so far, reducing repetition.
    \item \textbf{Presence Penalty}: Penalizes new tokens based on whether they appear in the text so far, encouraging the model to talk about new topics.
\end{itemize}

Beyond these parameters, other choices also influence output quality, especially \textbf{model selection} (for example, different model families) and \textbf{API type}.

It is useful to distinguish between \textbf{completions} and \textbf{responses}: completions are generally \textbf{stateless} (each call starts fresh unless you resend context), while responses can be \textbf{stateful}, with server-side conversation state that can simplify multi-turn interactions.

\subsection{Prompting Strategies}
This section covers common prompting strategies that improve model performance.
Calibrating specifically is a key part of the design process, as overspecifying or underspecifying the prompt can lead to suboptimal outputs.
Cutting redudancy and keeping a good structure permits to get better results, since the model follows literally the instructions 

The best strategy to get a good prompt is to iteratively test and refine it, starting with a simple version and gradually adding complexity as needed.

\subsubsection{Prompt Structure}
Defining a clear \textbf{scope} for the model is crucial to prevent it from going off-topic or generating irrelevant information.
To better handle complex tasks, it's also important to use \textbf{structure formatting} in the prompt, which can help guide the model's attention and improve the quality of the output.
An example is to separate instructions, input data and output format into distinct sections, using clear headings or bullet points.
\\
The suggested structure for a prompt is the following:
\begin{itemize}
    \item \textbf{Role (optional)}: Define the role or persona the model should adopt (for example, "You are a helpful assistant").
    This differetiates the \textbf{direct instruction} prompting strategy, where you simply provide clear, literal instructions for simple tasks, from the \textbf{role prompting} strategy, where you shape the tone, vocabulary and depth of the response.
    \item \textbf{Goal}: Clearly state in one sentence the overall goal or task (for example, "Summarize the following text in one sentence"). (This last part doesn't improve factual correctness)
    \item \textbf{Context}: Provide any necessary background information, data, or examples that the model needs to complete the task.
    \item \textbf{Constraints}: Specify any rules like audience, length limits and tone.
    \item \textbf{Output Format}: Define the expected format of the output (for example, "Return a JSON object with the following fields...").
    \item \textbf{Verify}: Ask the model to check its own output for errors or inconsistencies, which can help improve reliability.
\end{itemize}
Also \textbf{where you place instructions} relative to data significantly affects the quality, especially with long data.
The common rule is to place long data first, instructions last and query at the very end.

\subsubsection{Few-shot prompting}
Few-shot prompting involves providing the model with a few examples of the desired output format or behavior within the prompt itself.
This can help the model understand the task better and generate more accurate responses.
\\
It's important to choose diverse examples covering typical cases and edge cases, and keep a consistent format across examples to help the model learn the pattern.
Also ordering examples matters: place the most similar examples last, closer to the actual task, as the model tends to pay more attention to recent context.

\subsubsection{Chain-of-thought prompting}
LLMs generate tokens sequentially.
When you force intermediate reasoning steps, each subsequent token is conditioned on better context.
In other words, the model "thinks" by writing, and by prompting it to write down its reasoning process, you can often improve the quality of the final answer, especially for complex tasks that require multiple steps of reasoning.
\\
There are several variations of this strategy:
\begin{itemize}
    \item \textbf{Zero-shot CoT}: You simply ask the model to explain its reasoning process before giving the final answer, without providing any examples.
    \item \textbf{Few-shot CoT}: You provide examples of the reasoning process along with the final answer, which can help the model learn how to structure its reasoning.
    \item \textbf{Structured CoT}: You explicitly define the intermediate steps or sub-tasks that the model should follow, which can help guide its reasoning process more effectively.
\end{itemize}

%#TODO: Other slides remaining: from schema extraction and on